The curious world of constructive mathematics

### 2022-09-10 algar-lecture-notes.tex
More background on generic elements and models

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\FF}{\mathbb{F}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\CCC}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\let\acuteH\H
\newcommand{\konig}{K\acuteH onig}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\OOO}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\L}{\mathcal{L}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ccc}{\mathfrak{c}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\?}{\,{:}\,}
\renewcommand{\_}{\mathpunct{.}\,}

@@

\newcommand{\cl}{\mathrm{cl}}
\newcommand{\notnot}{\emph{not~not}}
\newcommand{\Sh}{\mathrm{Sh}}

@@

Toposes allow us to bring further entities into being, entities which do not
exist in the standard topos but which nevertheless affect the story of the
standard topos. We will focus on:
\begin{enumerate}
\item The \emph{generic ring} -- which cryptically is also a field
\item The \emph{generic prime filter}~$\ppp_\star$ of a given ring~$A$ -- which
XXX
\item The \emph{generic surjection}~$\NN \to X$ -- which exists also when~$X$
is uncountable
\end{enumerate}
The complex numbers can be compiled down to pairs of real numbers and hence
exhibit already in their very construction an intimate connection with the real
numbers; in a similar vein, these topos-theoretic generic models can be
compiled away in that proofs employing them can be unrolled, in a mechanical
fashion even, so as to not mention them.

@@

>>>>

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\FF}{\mathbb{F}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\CCC}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\EEE}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\let\acuteH\H
\newcommand{\konig}{K\acuteH onig}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\OOO}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\L}{\mathcal{L}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ccc}{\mathfrak{c}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\?}{\,{:}\,}
\renewcommand{\_}{\mathpunct{.}\,}

@@

\newcommand{\cl}{\mathrm{cl}}
\newcommand{\notnot}{\emph{not~not}}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\Set}{\mathrm{Set}}

@@

Toposes allow us to bring further entities into being, entities which do not
exist in the standard topos but which nevertheless affect the story of the
standard topos. We will focus on:
\begin{enumerate}
\item The \emph{generic ring} -- which cryptically is also a field
\item The \emph{generic prime filter} of a given ring~$A$ -- which
XXX
\item The \emph{generic surjection}~$\NN \to X$ -- which exists also when~$X$
is uncountable
\end{enumerate}
The complex numbers can be compiled down to pairs of real numbers and hence
exhibit already in their very construction an intimate connection with the real
numbers; in a similar vein, these topos-theoretic generic models can be
compiled away in that proofs employing them can be unrolled, in a mechanical
fashion even, so as to not mention them.

\section{Ring-theoretic prelude: the method of indeterminates}

The Cayley--Hamilton theorem of linear algebra states that every square matrix
over every commutative ring is a root of its characteristic polynomial. Among
its many proofs, the following one is particularly interesting from a logical
point of view: It is enough to verify the claim for the \emph{ring-theoretic
generic~$(n \times n)$-matrix}~$(A_{ij})_{ij}$ over the polynomial
ring~$\ZZ[A_{ij}\,|\,i,j\in\{1,\ldots,n\}]$, since this matrix can specialize
to every specific matrix over every commutative ring and this kind of specialization
preserves the~$n^2$ polynomial identities expressing that the matrix is a root
of its characteristic polynomial. Now to check this special case, it suffices to
verify it for every choice of complex numbers~$a_{ij} \in \CC$ in place of the
formal indeterminates~$A_{ij}$. For complex matrices, the claim follows from
the observation that the claim is immediate for the diagonalizable complex
matrices and that these are dense in the space of all complex matrices.

This approach is known as the \emph{method of indeterminates} and is amenable
to many situations. For instance, there is the
\emph{ring-theoretic generic~$n$-th root} (over~$\ZZ[X]/(X^n-1)$), the
\emph{ring-theoretic generic nilpotent element of index~$\leq n$}
(over~$\ZZ[X]/(X^n)$), the \emph{ring-theoretic generic power series}
(over~$\ZZ[A_0,A_1,\ldots]$), and many others. The central features of the
method of indeterminates are:
\begin{enumerate}
\renewcommand{\theenumi}{\arabic{enumi}}
\item We extend a certain base ring, often~$\ZZ$, to a larger ring~$R_0$ containing
the \emph{generic instance} of the problem at hand.
\item For every specific instance in every ring~$R$, there is a unique
homomorphism~$R_0 \to R$ mapping the generic instance to that one.
\item Since homomorphisms of rings preserve polynomial identities, every
polynomial identity (and in fact even every property which can be expressed as a
geometric formula in the language of rings) satisfied by the generic instance
passes down to every specific instance. In fact, the generic instance satisfies
exactly those polynomial identities which are provable from the ring axioms.
\item Crucially, the generic instance is applicable to special techniques (for
instance involving complex numbers) and has special properties which themselves
do not pass down to every specific instance, but which are useful in
establishing those that do.
\end{enumerate}

The generic models facilitated by topos theory generalize the ring-theoretic
generic elements. The latter are restricted to those kinds of elements which
can be described by polynomial identities (such as~$X^n - 1$ for~$n$-th roots).
For example, there is no ring-theoretic generic non-nilpotent element
or ring-theoretic generic regular element (a ring element~$x \in A$ is
\emph{regular} if and only if for all~$y \in A$, $xy = 0$ implies~$y = 0$).
The topos-theoretic generic models, in contrast, support arbitrary geometric
implications as defining properties, and are not restricted to mere elements
(or finite or infinite lists of elements): For instance, as elucidated in the
next sections, there is the topos-theoretic \emph{generic ring} or the
topos-theoretic \emph{generic prime filter} of any given ring.

Briefly, for every geometric theory~$\TT$ (set of geometric implications as
axioms) over any signature, there exists a topos-theoretic generic model
of~$\TT$. The analogy with the generic elements of ring theory is as follows:
\begin{enumerate}
\renewcommand{\theenumi}{\arabic{enumi}}
\item We extend the base universe, often the standard topos~$\Set$, to a larger
topos~$\Set[\TT]$ containing the \emph{generic model}~$U_\TT$ of~$\TT$.
\item For every specific~$\TT$-model~$M$ in any topos~$\EEE$, there is a unique
map~$\Set[\TT] \to \EEE$ mapping~$U_\TT$ to~$M$.\footnote{To be more precise,
the category of maps~$\mathrm{Hom}(\Set[\TT],\EEE)$ is equivalent to the
category of~$\TT$-models in~$\EEE$, naturally in~$\EEE$. By ``map'', we mean
what is called the ``inverse image part of geometric morphisms'' in topos
theory.}
\item Since such maps preserve geometric implications, every property
which can be expressed as a geometric implication passes from~$U_\TT$ to every
specific model. In fact, the generic model satisfies exactly those geometric
implications which are provable from the axioms of~$\TT$.
\item Crucially, the generic model is applicable to special techniques and has
special properties which themselves do not pass down to every specific model,
but which are useful in establishing those that do.
\end{enumerate}

@@

\begin{exercise}[Limitations of ring-theoretic generics]
Show that each of the following hypothetical ring-theoretic generics do not
exist:
\begin{enumerate}
\item the generic nilpotent element (of arbitrary index)
\item the generic regular element
\item the generic prime element
\end{enumerate}
That is, show that there is no ring~$R_0$ containing an element~$x_0$ of the
desired kind such that for every such element~$x$ of every ring~$R$, there is a
unique homomorphism~$R_0 \to R$ mapping~$x_0$ to~$x$.
\end{exercise}
